Efficient spherical Convolutional Neural Networks with Healpix sampling for cosmological applications

### 2018-05-11 scnn.tex
small corrections

\todo{
\subsection{Healpix sampling}
Should we have this subsection to explain the advantages of Healpix?
\begin{itemize}
	\item Define the gedesic grid?
	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
\end{itemize}
}

@@

\subsection{Spherical convolutions} 
The major challenge to generalize CNN to a
spherical domain is to generalize convolution. In the context of CNN, we found
two particular approaches to address this issue.

@@

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere. \nati{Michael: Under some very
specific hypothesis, this second approach is a particular case of our method,
i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
on a grid == 2d convolution. Do you think we should mention that? I think we
should not.}

\todo{Check this unpublished work \cite{khasanova2017graph}}

@@

@@

\subsection{Graph creation}
In the classical 2-dimenstional convolution, each pixel is connected with the
same weight to its 4 closest neighbors. We use a weighted undirected graph to
generalize this idea to the ubiquitous HealPix
sampling~\citep{gorski2005healpix}.  Each pixel is represented by a node
(vertex) connected to his $8$ or $7$ closest neighbors.\footnote{For some
pixels, the $8^{th}$ nearest neighbor is not well defined.} Given the set of
nearest neighbors, we define the weight matrix $W$ using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$
on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected
pixels $i$ and $j$. This weighting scheme is important as the distances between
pixels varries du to the small sampling irregularities. Some of the
irregularities have been highlighted in the graph displayed in
Figure~\ref{fig:healpix_graph_4}.

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.45\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using a nside parameter of $4$.
Nodes with 7 neighboors are highlighted.
}
\label{fig:healpix_graph_4}
\end{figure}

@@

Because of the irregular sampling, there is no obvious way to define convolution
directly from the pixel values. Hence we first define a Fourier transform on the
graph.

Following~\cite{shuman2013emerging}, the graph normalized graph Laplacian
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$ is a second order differential operator
that can be used to define the graph Fourier basis. Here $D$ is the diagonal
matrix where $\D_{ii}=\b{d}_i$. By construction the Laplacian is symetric postive
semi-definite and hence can be decomposed as $\L=\U \bLambda \U^*$, where $U$ is an
orthonormal matrix of eigenvectors and $\bLambda$, the diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors.
The graph Fourier transform of $f$ is simply its projection on $U$ given by 
$\hat{\b{f}}=\U^*\b{f}$. Similarly the inverse graph Fourier transform reads $\b{f}=\U\hat{\b{f}}$.
Note that instead of frequencies, the Fourier modes are crescently ordered with
respect of the Laplacian eigenvalues $\bLambda$. In a sense, the Laplacian
eigenvalues correspond to the squared frequencies.

@@

As shown in Figure~\ref{fig:graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics ressembles the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling.

@@

The major difference with classical convolution is that the convolution kernel
$h$ cannot be thought as $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization puposes, one can
look at effect of the convolution on a dirac, i.e. one column of the matrix
$h(\L)$. \nati{However, due to the non regularity of the graph (i.e. the fact
that there is not perfect sampling on the sphere), this visualization will
differ from one node to another. In the specific case of the full sphere, these
difference are negligible in most of the cases. When considering only subpart of
the sphere, one will observe important border effects.}

\paragraph{Example: heat diffusion}
Let us consider the heat diffusion problem 
\begin{equation} \label{eq:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition 
$\b{f}(0)$, the solution of~\ref{eq:heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0), 
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
correspond to the squared frequencies and hence $K_t$ can be considered a
Gaussian kernel. In Figure~\ref{fig:gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various 
time $t$. In  Figure~\ref{fig:gaussian_filters_visualization} of
Appendix~\ref{app:filter_visualization}, we show different visualizaton of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(Figure~\ref{fig:gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parametrization, well approximate it.

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/gaussian_filters_sphere.pdf}
\caption{Convolution comparizon (Nside=16). 
Top: Diffusion of a unit of heat for different time $t$ using the graph. 
Bottom: spherical symmetric Gaussian smoothing for different $\sigma$ (arcmin). 
Relative difference between graph convolution and spherical smoothing: $10.4$\%, $5.7$\%, $4.8$\%, $3.8$\% }
\label{fig:gaussian_filters_comparizon}
\end{figure*}

It is another sign that the constructed graph is able to capture the spherical
structure of the healpix sampling. Furthemore, in some application, where the
exactitude of the convolution is not requirement, such as de-noising, it means
that we could use graph convolution instead. Depending on the convolution kernel
this may reveal much more efficient. In the case of a neural network, the fact
that both convolution are not identical is not very relevant since the network
will adapt the convolution at use.

@@

\subsection{Coarsening}
The HealPix sampling present the advantage to have a natural corsening method
based on its construction. Each nside level divides each pixel into 4 subpixels.
We define the pooling process as the inverse operation grouping the 4 subpixels.
Figure~\ref{fig:pooling} illustrates the process.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
\caption{Pooling illustration for 2 levels using some part of the the mollview projection. 
The final patch represent 1/12 of the sphere.}
\label{fig:pooling}
\end{figure}

@@

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the usefulness of the spherical neural network
on a discrimination problem. Our implementation is availlable as small python package at \url{https://our.code.link.com}. 
Furthermore, all the data to reproduce our experiments are available online.\footnote{\url{https://doi.org/10.5281/zenodo.????} \todo{correct DOI}}

@@

As a preprocssing step, we first remove the mean of each sample and apply a
spherical gaussian smoothing kernel with a radius of $1$ arcmin.

@@

Leveraging the properties of the HealPix sampling, we can split each simulation
into $12*4^o$ ($o=0,1,2,\dots$) samples that span a smaller part of the sphere.
As shown in Figure~\ref{fig:subpart_sphere}, we used $o=0,1,2$ as the resulting
samples are large enough to suffer from the effet of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

We would like to build a robust classifier, i.e. an automatic way to
discriminate between the two classes with the presence of noise. For all the
eperiment, we work with zero mean Gaussian noise. \todo{Tomek: could you put a
justification.} All classifiers are trained with noisy samples. To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced at each iterations. We trained our
model between $5$ and $40$ epochs depending of the complexity and the size of
the problem. The testing set was simply used to assess the global performance of
the network and if the it was overfitting. 
Training took between $1$ and $10$ hours using a Nvidia 1080 TI.

We compare the spherical CNN with a handcrafted feature classifier. \cite{...}
\todo{Tomek: we need probably 1 or two other classifiers.} The features are
constructed using a normalized histogram of the values. We then train a linear
SVN classifier. We tried different kernels, but they where not leading to better
results. For a fair comparizon, we increase the amount of training data of the
SVN classifier until the testing error plateau to fix value.

@@

We present in Figure~\ref{fig:results} the results for 5 different noise levels and $3$ different data sizes. The spherical neural network is always significantly better than the hand-crafted classfier.

@@

In Figure~\ref{fig:gaussian_filters_visualization}, we present three different
ways to plot the graph convolution kernel $K_t(x)=e^{-\tau t x}$.  First, it can
be ploted in the graph spectral domain. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. Second, we can convolve a
delta on the sphere and plot a gnomonic view of the results. Third, since we are
working on the sphere, we can only plot the section of the convoluted delta. In
Figure~\ref{fig:index_section}, we display the selected indexes for the section
plotting. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are likely to change depending on the
chosen delta.

>>>>

\todo{
\subsection{Healpix sampling}
Should we have this subsection to explain the advantages of Healpix?
\begin{itemize}
	\item Define the spherical grid?
	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
\end{itemize}
}

@@

\subsection{Spherical convolutions} 
The major challenge to define Convolution Neural Network (CNN) onto a
spherical domain is to handle a convolution suitable for this irregular domain. 
We found two particular approaches to address this issue.

@@

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere. 
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}

@@

We are not the first one to leverage a graph to define a CNN on a sphere.
\cite{khasanova2017graph} use the same idea to handle omnidirectional images.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as \cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.

@@

\subsection{Graph creation}
In the classical 2-dimenstional convolution, each pixel is connected with the
same weight to its 4 closest neighbors. We use a weighted undirected graph to
generalize this idea to the ubiquitous HealPix
sampling~\citep{gorski2005healpix}.  Each pixel is represented by a node
(vertex) connected to his $8$ or $7$ closest neighbors.\footnote{For some
pixels, the $8^{th}$ nearest neighbor is not well defined.} Given the set of
nearest neighbors, we define the weight matrix $W$ using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$
on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected
pixels $i$ and $j$. This weighting scheme is important as the distances between
pixels varies due to the small sampling irregularities. Some of the
irregularities have been highlighted in the graph displayed in
Figure~\ref{fig:healpix_graph_4}.

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.45\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using an nside parameter of $4$.
Nodes with 7 neighbors are highlighted.
}
\label{fig:healpix_graph_4}
\end{figure}

@@

Because of the irregular sampling, there is no obvious way to define the convolution
directly from the pixel values. Hence we first define a Fourier transform on the
graph.

Following~\cite{shuman2013emerging}, the graph normalized graph Laplacian
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$ is a second order differential operator
that can be used to define the graph Fourier basis. Here $D$ is the diagonal
matrix where $\D_{ii}=\b{d}_i$. By construction the Laplacian is symetric positive
semi-definite and hence can be decomposed as $\L=\U \bLambda \U^*$, where $U$ is an
orthonormal matrix of eigenvectors and $\bLambda$, the diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors.
The graph Fourier transform of $f$ is simply its projection on $U$ given by 
$\hat{\b{f}}=\U^*\b{f}$. Similarly the inverse graph Fourier transform reads $\b{f}=\U\hat{\b{f}}$.
Note that instead of frequencies, the Fourier modes are crescently ordered with
respect of the Laplacian eigenvalues $\bLambda$. In a sense, the Laplacian
eigenvalues correspond to the squared frequencies.

@@

As shown in Figure~\ref{fig:graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling.

@@

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker, i.e. one column of the matrix
$h(\L)$. \nati{However, due to the non-regularity of the graph (i.e. the fact
that there is not perfect sampling on the sphere), this visualization will
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. When considering only subpart of
the sphere, one will observe important border effects.}

\paragraph{Example: heat diffusion}
Let us consider the heat diffusion problem 
\begin{equation} \label{eq:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition 
$\b{f}(0)$, the solution of~\ref{eq:heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0), 
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
correspond to the squared frequencies and hence $K_t$ can be considered a
Gaussian kernel. In Figure~\ref{fig:gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various 
time $t$. In  Figure~\ref{fig:gaussian_filters_visualization} of
Appendix~\ref{app:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(Figure~\ref{fig:gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/gaussian_filters_sphere.pdf}
\caption{Convolution comparison (Nside=16). 
Top: Diffusion of a unit of heat for different times $t$ using the graph. 
Bottom: spherical symmetric Gaussian smoothing for different $\sigma$ (arcmin). 
Relative difference between graph convolution and spherical smoothing: $10.4$\%, $5.7$\%, $4.8$\%, $3.8$\% }
\label{fig:gaussian_filters_comparizon}
\end{figure*}

It is another sign that the constructed graph is able to capture the spherical
structure of the HealPix sampling. Furthermore, in some application, where the
exactitude of the convolution is not a requirement, such as de-noising, it means
that we could use the graph convolution instead. Depending on the convolution kernel
this may reveal much more efficient. In the case of a neural network, the fact
that both convolution are not identical is not very relevant since the network
will adapt the convolution at use.

@@

\subsection{Coarsening}
The HealPix sampling presents the advantage to have a natural coarsening method
based on its construction. Each Nside level divides each pixel into 4 subpixels.
We define the pooling process as the inverse operation grouping the 4 subpixels.
Figure~\ref{fig:pooling} illustrates the process.
\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
\caption{Pooling illustration for 2 levels using some part of the mollview projection. 
The final patch represents 1/12 of the sphere.}
\label{fig:pooling}
\end{figure}

@@

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the usefulness of the spherical neural network
on a discrimination problem. Our implementation is available as a small python package at \url{https://our.code.link.com}. 
Furthermore, all the data to reproduce our experiments are available online.\footnote{\url{https://doi.org/10.5281/zenodo.????} \todo{correct DOI}}

@@

As a pre-processing step, we first remove the mean of each sample and we apply a
spherical Gaussian smoothing kernel with a radius of $1$ arcmin.

@@

Leveraging the properties of the HealPix sampling, we can split each simulation
into $12*4^o$ ($o=0,1,2,\dots$) samples that span a smaller part of the sphere.
As shown in Figure~\ref{fig:subpart_sphere}, we used $o=0,1,2$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

@@

We would like to build a robust classifier, i.e. an automatic way to
discriminate between the two classes with the presence of noise. For all the
experiments, we work with zero mean Gaussian noise. \todo{Tomek: could you put a
justification.} All classifiers are trained with noisy samples. To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced at each iteration. We trained our
model between $5$ and $40$ epochs depending on the complexity and the size of
the problem. The testing set was simply used to assess the global performance of
the network and to assess if were overfitting. 
Training took between $1$ and $10$ hours using a Nvidia 1080 TI.

We compare the spherical CNN with a handcrafted feature classifier. \cite{...}
\todo{Tomek: we probably need 1 or two other classifiers.} The features are
constructed using a normalized histogram of the values. We then train a linear
SVN classifier. We tried different kernels, but they were not leading to better
results. For a fair comparison, we increase the amount of training data of the
SVN classifier until the testing error plateau to fix value.

@@

We present in Figure~\ref{fig:results} the results for 5 different noise levels and $3$ different data sizes. The spherical neural network is always significantly better than the hand-crafted classifier.

@@

In Figure~\ref{fig:gaussian_filters_visualization}, we present three different
ways to plot the graph convolution kernel $K_t(x)=e^{-\tau t x}$.  First, they can
be plotted in the graph spectral domain. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. Second, we can convolve a
delta on the sphere and plot a gnomonic view of the results. Third, since we are
working on the sphere, we can only plot the section of the convoluted delta. In
Figure~\ref{fig:index_section}, we display the selected indexes for the section
plotting. Note that, because of the small irregularities in the HealPix
sampling, the second and the third methods are likely to change depending on the
chosen delta.
